---
title: "Creating a FacileDataSet from ExpressionPlot"
author: "Jonathan Carroll, Vincent Rouilly, Peter Haverty, and the FacileExplorer Working Group"
date: "Last Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    code_folding: show
    keep_md: true
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

required_packages <- c("desc", "devtools", "glue")
for (pkg in required_packages) {
  stopifnot(requireNamespace(pkg, quietly = TRUE))
}
```


# Overview

The current workflow for ingesting data into `FacileExplorer` is to create a `Facile___DataSet` R package. This allows both the data and some complementary R scripts to be contained together. For now we will focus on just the data.

The path for now is to go via a `list` of harmonized and annotated `ExpressionSet` or `SummarizedExperiment` objects. We have broken the steps down into documented chunks here.

This conversion is based on a WIP script found here: https://stash.intranet.roche.com/stash/projects/FACILEDATA/repos/faciledataset/browse/inst/scripts/gcell_rnaseq_as.FacileDataSet.R?at=refs%2Fheads%2Ffeature%2Fas.FacileDataSet

The result of this workflow should be a Facile___DataSet. Let's start by filling in that gap

# Define the Facile___DataSet

The Facile___DataSet needs a name (in the ___ component -- this structure is mandatory) and a version

```{r define}
dataset_name <- "GCell"
FDS_name <- glue::glue("Facile{dataset_name}DataSet")
FDS_version <- "0.1.2"
```

This will therefore produce a `r FDS_name` package.

# Install the `FacileData` R Package

available from ssh://git@stash.intranet.roche.com:7999/faciledata/faciledataset.git

```{r}
library(FacileData)
```

## Note

- This script requires the Genentech bundled R 3.4.3 distribution. The R 3.5.0 distribution is not able to detect the `ExpressionPlot` projects used and complains about `S4Vectors`.
- PRJ0013166 (NGS114) has been superseded and this script will now use PRJ0013167.
- PRJ0013155 (NGS171) has been superseded and this script will now use PRJ0013730 (not sure if connected at all).
- `ExpressionPlot` R package requires that the calls be made on rescomp, since it uses an internal perl script.
- If building `FacileData` from source, make sure to document the functions (e.g. `devtools::document` or via Build Tools) as this is not always complete in the repo, and functionality will not be exported otherwise.

# Load data from `ExpressionPlot`

Load two datasets directly from `ExpressionPlot`. These have been superseded, so we will skip the warnings.

```{r load ep data}
library(ExpressionPlot)

ngs114 <- ep.project.from.id("PRJ0013166", standard.checks = FALSE)
ngs171 <- ep.project.from.id("PRJ0013155", standard.checks = FALSE)

ds <- list(
  NGS114 = ep.SummarizedExperiment(ngs114, attach.annot = TRUE),
  NGS171 = ep.SummarizedExperiment(ngs171, attach.annot = TRUE)
)
```

# Subset and rename

The feature space and phenoData spaces need to be harmonized between all datasets within a `Facile___DataSet`. The following chunk:

- prefixes the GeneIDs
- subsets the phenoData to a limited set of columns, and renames them
- populates the metadata with definitions of the phenoData columns

For now, categorical covariates are the more common variables as there are few tools available yet to deal with continuous covariates.

```{r subset phenotypes}
## examples of type: data_batch, clinical, tumor_classification, response, IHC, mutation, safety, conmeds
## examples of class: numerical, categorical, right_censored
## (**can't create new type or class**)
ds <- lapply(ds,
             function(x) {
               rownames(x) = paste0("GeneID:",rownames(x))
               f = colnames(mcols(x)) = c("aliases", "effective_length", "feature_type", "name", "meta")
               mcols(x)$source = "IGIS"
               mcols(x)$feature_type = "entrez"
               mcols(x)$feature_id = rownames(x)
               colData(x) = colData(x)[,c("PRIMARY_TISSUE", "GENDER", "TISSUE_DIAGNOSIS", "ETHNICITY", "TISSUE_METACLASS_ONCOLOGY")]
               colnames(colData(x)) = c("primary_tissue", "gender", "diagnosis", "ethnicity", "tissue")
               metadata(colData(x)) = list(
                 primary_tissue = list(label = "Primary Tissue",
                                       description = "Primary tissue source",
                                       type = "tumor_classification",
                                       class = "categorical"
                 ),
                 gender = list(label = "Gender/Sex",
                               description = "In the 'ratio between chrX:chrY' sense.",
                               type = "clinical",
                               class = "categorical"
                 ),
                 diagnosis = list(label = "Tissue Diagnosis",
                                  description = "Tissue metaclass oncology",
                                  type = "tumor_classification",
                                  class = "categorical"
                 ),
                 ethnicity = list(label = "Ethnicity",
                                  description = "Patient ethnicity",
                                  type = "clinical",
                                  class = "categorical"
                 ),
                 tissue = list(label = "Tissue Group",
                               description = "Rollup of tissue type to defined vocab",
                               type = "tumor_classification",
                               class = "categorical"
                 )
               )
               x
             })
```

The overall metadata for the studies can also be populated

```{r metadata}
metadata(ds[[1]]) <- list(url = "http://gene.com", description = "This is NGS114")
metadata(ds[[2]]) <- list(url = "http://gene.com", description = "This is NGS171")
```

# Convert to `FacileDataSet`

The `Facile___DataSet` will be stored in a directory (specified, should not exist yet). The contents of this new directory need to be copied into a new package directory, which we will also create.

```{r create FDS}
DIR <- file.path("../../FacileVerse", FDS_name)
unlink(DIR, recursive = TRUE, force = TRUE)
devtools::create(DIR) ## create a package
## start an installation directory
## the parent directory of the 
## FacileDataSet must already exist.
dir.create(file.path(DIR, "inst", "extdata"), recursive = TRUE) 

gcell_fds <- as.FacileDataSet(ds,
                              path = file.path(DIR, "inst", "extdata", FDS_name),
                              assay_name = "rnaseq",
                              assay_type = "rnaseq",
                              source_assay = "counts",
                              dataset_name = "gCell",
                              organism = "Homo sapiens"
)
```

# Try it out

Before we package the dataset entirely, we can reassure ourselves that it works as expected using the `FacileDataSet` API

```{r testing}
library(dplyr)
library(reshape2)

samples <- sample_info_tbl(gcell_fds) %>% collect
samples <- fetch_sample_covariates(gcell_fds, samples, "tissue") %>%
  filter(variable == "tissue" & value == "Breast") %>%
  collect
genes <- c(ERBB2 = 'GeneID:2064', GRB7 = 'GeneID:2886')
exprs <- gcell_fds %>% fetch_assay_data(genes, samples, 'rnaseq', normalized = TRUE)
ew <- exprs %>% dcast(dataset + sample_id ~ feature_name, value.var = 'value')
cor(ew[,"ERBB2"], ew[,"GRB7"]) # ~ 0.76

library(ggplot2)
theme_set(theme_bw())
ggplot(ew, aes(ERBB2, GRB7)) +
  geom_point()
```

# Create the accessor function

`FacileExplorer` requires that the Facile___DataSet be accessible via a named function, so we create it in the R/ directory

```{r accessor}
writeLines(
  glue::glue("##' A connection to the {FDS_name}",
             "#'",
             "#' @export",
             "#' @return A \\code{{FDS_name}} object",
             "{FDS_name} <- function(path, cache_size=80000,",
             "                               db.loc = c('reference', 'temporary', 'memory')) {{",
             "  if (missing(path) || is.null(path)) {{",
             "    path <- system.file('extdata', '{FDS_name}', package='{FDS_name}')",
             "  }}",
             "  db.loc <- match.arg(db.loc)",
             "  FacileDataSet(path, cache_size = cache_size, db.loc = db.loc)",
             "}}", .sep = "\n"),
  file.path(DIR, "R", glue::glue("{FDS_name}.R"))
)
```

# Build the Facile___DataSet package

Building the package loads it into your environment. We will also save a version for export.

```{r build}
## custom-annotation directory would be removed if empty
writeLines("", file.path(DIR, "inst", "extdata", FDS_name, "custom-annotation", "README.md"))
## update package version
desc::desc_set_version(FDS_version, file = file.path(DIR, "DESCRIPTION"))
## document, install, and build source tar.gz
devtools::document(DIR)
devtools::install(DIR)
devtools::build(pkg = DIR, path = "../../FacileVerse/")
```